Data on Death Registrations Illustrates Issues with Ontario's Records Maintenance

### 2023-02-02 outputs/paper/paper.qmd
updated re-organized scripts, fixed import of data to quarto file,

#### Workspace Set-up ####

#### install.packages(tidyverse) ####
#### install.packages(janitor) ####
#### install.packages(ggplot2) ####
#### install.packages("knitr") ####
#### install.packages("RColorBrewer") ####
#### install.packages("kableExtra") ####
#### install.packages("bookdown") ####
#### install.packages("rmarkdown") ####
#### install.packages("patchwork") ####

library(tidyverse)
library(opendatatoronto)
library(janitor)
library(ggplot2)
library(dplyr)
library(knitr)
library(RColorBrewer)
library(kableExtra)
library(bookdown)
library(patchwork)

@@

Analysis for this paper utilizes R the statistical programming language [@citeR], along with several packages. These packages are, tidyverse [@citetidyverse], janitor [@citejanitor], and dplyr [@citedplyr]. The data I am analyzing come from Open Data Toronto and it is imported using the opendatatoronto package [@citeopendatatoronto]. **All figures in this paper have been created using ggplot2 [@citeggplot2] and the tables have been created with knitr [@citeknitr] and kableExtra [@citekablextra], packages. The color styles in my graph were created by using the RColorBrewer [@citercolorbrewer] package and any graph combinations were made using the Patchwork [@citepatchwork] package.**

@@

```{r}
#| include: FALSE

#### Read in Data ####

raw_death_registry_data <-
  read_csv(
    file = 
      "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/cba07a90-984b-42d2-9131-701c8c2a9788/resource/cc418325-b3d0-4afd-ae64-ccb9b71c549c/download/Death%20Registry%20Statistics%20Data.csv",
    show_col_types = FALSE,
    skip = 0
  )

#### Creating a CSV File for Raw Data ####
write_csv(
  x = raw_death_registry_data,
  file = "death_registry.csv" #create raw data .csv file and store it in inputs folder
)

#### Data Cleaning ####

#### Clean Names ####
cleaned_death_registry_data <-
  clean_names(raw_death_registry_data)

#### Simplifying Date #### 
cleaned_death_registry_data <-
  cleaned_death_registry_data |>
  separate(
    col = time_period,
    into = c("Year", "Month"),
    sep = "-"
  ) |>
  select(civic_centre, death_licenses, place_of_death, Year, Month)

#### Removing 2011-2015 and 2021-2023 Data for Relevance ####
cleaned_death_registry_data <- cleaned_death_registry_data[-c(1:371,746:847), ]

#### Recoding Months to be More Meaningful ####
cleaned_death_registry_data <- cleaned_death_registry_data |> 
  mutate( 
    Month = 
      recode( 
        Month, #abbreviated months to better fit graphs withotu losing meaning
        "01" = "Jan", 
        "02" = "Feb", 
        "03" = "Mar",
        "04" = "Apr",
        "05" = "May",
        "06" = "Jun",
        "07" = "Jul",
        "08" = "Aug",
        "09" = "Sep",
        "10" = "Oct",
        "11" = "Nov",
        "12" = "Dec"
      ) 
  )

#### Recoding Civic Centres to be More Meaningful ####
cleaned_death_registry_data <- cleaned_death_registry_data |> 
  mutate( 
    civic_centre = 
      recode( 
        civic_centre, 
        "ET" = "Etobicoke", 
        "NY" = "North York", 
        "SC" = "Scarborough", 
      ) 
  )

#### Ordering Months #####
cleaned_death_registry_data$Month <- factor(cleaned_death_registry_data$Month, levels=c("Jan", "Feb","Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

#### Filtering Out "Toronto" Civic Centre from Data Due to Missing Data ####
cleaned_death_registry_data <- cleaned_death_registry_data |>
  filter(!grepl('TO', civic_centre))

#### Creating a CSV File for Cleaned Data ####
write_csv(
  x = cleaned_death_registry_data,
  file = "cleaned_death_registry.csv"  #create cleaned data .csv file and store it in inputs folder
)

([@fig-deathlicensesbymonthandyear])

@@

@@

month_by_death_license <- ggplot(data=cleaned_death_registry_data, aes(x = Month, y = death_licenses, fill = civic_centre)) + #assign ggplot to new variable in order to use patchwork package later
  geom_bar(stat = "identity") +
  theme_classic() + 
  labs(
    x = "Month",
    y = "Death Licenses",
    fill = "Civic Centre"
  ) +
  scale_fill_brewer(palette="Set2")+
  theme(legend.position = "none") + #remove legend as next graph will have it for better readability
    theme(axis.title.y = element_text(size = 10))+
    theme(axis.title.x = element_text(size = 10))

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-avgnumofdeathsmonthet
#| tbl-cap: "Average Number of Deaths Per Month from 2016-2020 at Etobicoke Centre"

@@

et_all_avg_num_deaths_by_month_data|>
  select(Month, "Average # of Deaths Registered.y","Average # of Deaths Registered.x.x", "Average # of Deaths Registered.y.y", "Average # of Deaths Registered.x.x.x", "Average # of Deaths Registered.y.y.y")|> #previous step created a duplicate column, use select() to remove the second column in dataset
     kable(
    col.names = c( #rename columns to be more meaningful
      "Month",
      "2016",
      "2017",
      "2018",
      "2019",
      "2020"
    ),
    digits = 0,
    booktabs = TRUE,
    linesep = "") |>
    add_header_above(c(" " = 1, "Average Number of Deaths" = 5)) |>
  kable_styling(full_width = TRUE)

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-avgnumofdeathsmonthsc
#| tbl-cap: "Average Number of Deaths Per Month from 2016-2020 at Scarborough"

@@

sc_all_avg_num_deaths_by_month_data|>
  select(Month, "Average # of Deaths Registered.y","Average # of Deaths Registered.x.x", "Average # of Deaths Registered.y.y", "Average # of Deaths Registered.x.x.x", "Average # of Deaths Registered.y.y.y")|> #previous step created a duplicate column, use select() to remove the second column in dataset
  kable(
    col.names = c( #rename columns to be more meaningful
      "Month",
      "2016",
      "2017",
      "2018",
      "2019",
      "2020"
    ),
    digits = 0,
    booktabs = TRUE,
    linesep = "") |>
  add_header_above(c(" " = 1, "Average Number of Deaths" = 5)) |>
  kable_styling(full_width = TRUE)

@@

ny_all_avg_num_deaths_by_month_data|>
  select(Month, "Average # of Deaths Registered.y","Average # of Deaths Registered.x.x", "Average # of Deaths Registered.y.y", "Average # of Deaths Registered.x.x.x", "Average # of Deaths Registered.y.y.y")|> #previous step created a duplicate column, use select() to remove the second column in dataset
  kable(
    col.names = c( #rename columns to be more meaningful
      "Month",
      "2016",
      "2017",
      "2018",
      "2019",
      "2020"
    ),
    digits = 0,
    booktabs = TRUE,
    linesep = "") |>
  add_header_above(c(" " = 1, "Average Number of Deaths" = 5)) |>
  kable_styling(full_width = TRUE)
```

>>>>

#### workspace set-up ####

library(tidyverse)
library(opendatatoronto)
library(janitor)
library(ggplot2)
library(dplyr)
library(knitr)
library(RColorBrewer)
library(kableExtra)
library(bookdown)
library(patchwork)
library(readr)
library(here)

#### import cleaned_death_registry.csv ####
cleaned_death_registry_data <- read.csv(here::here("/Users/chloethierstein/Desktop/death_registry_analysis/inputs/data/cleaned_death_registry.csv"))

#### turning years into factor ####
cleaned_death_registry_data$Year <- as.factor(cleaned_death_registry_data$Year)
```

@@

Analysis for this paper utilizes R the statistical programming language [@citeR], along with several packages. These packages are, tidyverse [@citetidyverse], janitor [@citejanitor], and dplyr [@citedplyr]. The data I am analyzing come from Open Data Toronto and it is imported using the opendatatoronto package [@citeopendatatoronto]. **

All figures in this paper have been created using ggplot2 [@citeggplot2] and the tables have been created with knitr [@citeknitr] and kableExtra [@citekablextra], packages. The color styles in my graph were created by using the RColorBrewer [@citercolorbrewer] package and any graph combinations were made using the Patchwork [@citepatchwork] package.**

@@

(@fig-deathlicensesbymonthandyear)

@@

```{r}
#| include: false
#| message: false

@@

month_by_death_license <- ggplot(data=cleaned_death_registry_data, aes(x = Month, y = death_licenses, fill = civic_centre)) + #assign ggplot to new variable in order to use patchwork package later
  geom_bar(stat = "identity") +
  theme_classic() + 
  labs(
    x = "Month",
    y = "Death Licenses",
    fill = "Civic Centre"
  ) +
  scale_fill_brewer(palette="Set2")+
  theme(legend.position = "none") + #include legend at bottom to ensure readability when using patchwork package
  theme(axis.title.y = element_text(size = 10)) +
  theme(axis.title.x = element_text(size = 10))

```{r}
#| include: false
#| message: false

@@

```{r}
#| warning: false
#| echo: false
#| label: tbl-avgnumofdeathsmonthet
#| tbl-cap: "Average Number of Deaths Per Month from 2016-2020 at Etobicoke Centre"

@@

et_all_avg_num_deaths_by_month_data |>
  select(
    Month,
    "Average # of Deaths Registered.y",
    "Average # of Deaths Registered.x.x",
    "Average # of Deaths Registered.y.y",
    "Average # of Deaths Registered.x.x.x",
    "Average # of Deaths Registered.y.y.y"
  ) |> #previous step created a duplicate column, use select() to remove the second column in dataset
  kable(
    col.names = c(#rename columns to be more meaningful
      "Month",
      "2016",
      "2017",
      "2018",
      "2019",
      "2020"),
    digits = 0,
    booktabs = TRUE,
    linesep = ""
  ) |>
  add_header_above(c(" " = 1, "Average Number of Deaths" = 5))

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-avgnumofdeathsmonthsc
#| tbl-cap: "Average Number of Deaths Per Month from 2016-2020 at Scarborough Centre"

@@

sc_all_avg_num_deaths_by_month_data|>
  select(Month, "Average # of Deaths Registered.y","Average # of Deaths Registered.x.x", "Average # of Deaths Registered.y.y", "Average # of Deaths Registered.x.x.x", "Average # of Deaths Registered.y.y.y")|> #previous step created a duplicate column, use select() to remove the second column in dataset
  kable(
    col.names = c( #rename columns to be more meaningful
      "Month",
      "2016",
      "2017",
      "2018",
      "2019",
      "2020"
    ),
    digits = 0,
    booktabs = TRUE,
    linesep = "") |>
  add_header_above(c(" " = 1, "Average Number of Deaths" = 5))

@@

ny_all_avg_num_deaths_by_month_data|>
  select(Month, "Average # of Deaths Registered.y","Average # of Deaths Registered.x.x", "Average # of Deaths Registered.y.y", "Average # of Deaths Registered.x.x.x", "Average # of Deaths Registered.y.y.y")|> #previous step created a duplicate column, use select() to remove the second column in dataset
  kable(
    col.names = c( #rename columns to be more meaningful
      "Month",
      "2016",
      "2017",
      "2018",
      "2019",
      "2020"
    ),
    digits = 0,
    booktabs = TRUE,
    linesep = "") |>
  add_header_above(c(" " = 1, "Average Number of Deaths" = 5)) 
```

\clearpage
